### 2015-04-28 papers/histogram/thesis/thesis.tex

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased (i.e. Metropolis) Monte Carlo
simulations are incapable of sampling the low energy fluid states
which dominate the partition function at low (i.e. liquid-state)
temperatures. Several generic Monte Carlo ``histogram methods'' for
collecting statistics on low energy system states have been developed
\fixme{by whom?}, but little or no literature exists on their
systematic comparison and their application to the square-well
fluid. The square-well fluid in particular introduces application
challenges not manifest in traditional models for testing and
benchmarking such numerical techniques (e.g. the Ising model). We
propose to implement these methods in Monte Carlo simulations of the
square-well fluid, determine appropriate performance metrics for their
comparison, and identify those methods which are most efficient and
effective at sampling otherwise improbable system states.

@@

\fixme{talk more about phase transitions and the critical point?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the boundary between its liquid and
gaseous phases. The square-well fluid (Section \ref{sec:sw_fluid}) is
the simplest system with a liquid-vapor phase transition, and is
therefore of interest for studying the critical point of fluids. For a
generic square-well fluid to be observed in a liquid state, it must
have a localized distribution of the spheres which make up the
fluid. Such a distribution implies a low energy fluid state, as all
interactions between spheres are attractive. Conversely, a low energy
liquid state of a square-well fluid otherwise observable in both
liquid and gaseous phases implies a localized distribution of
spheres. The search for liquid states of the square-well fluid is thus
equivalent to the search for low energy states.

Monte Carlo simulations are a standard means for studying the
thermodynamic properties of a system. In the most straightforward
implementation of Monte Carlo fluid simulations (Section
\ref{sec:mc_sim}), however, highly localized distributions of fluid
elements are extremely unlikely to occur, making such simulations
impractical for studying the critical point of the square-well
fluid. Several generic computational methods, called histogram methods
(Section \ref{sec:histogram_methods}), exist for dealing with such a
problem; namely, that interesting regions of a system's state space
cannot be sampled via standard Monte Carlo methods in a reasonable
amount of time. There is no consensus, however, on the relative
efficacy of these methods. Furthermore, some methods, such as the
Wang-Landau method\cite{wang_landau}, require tuning several free
parameters that lack both ``obvious'' values and systematic means for
assigning them. This work will thus attempt to quantify and compare
the effectiveness of a few such methods.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased, Metropolis Monte
Carlo fluid simulations. Such an algorithm is ``unbiased'' in the
sense that it collects statistics (i.e. data) on all valid system
configurations with equal probability. Statistics whose collection
time scales as $\mathcal O\p{1}$ with system size, meaning that
increasing the number of spheres $N$ in the simulation does not affect
collection time, can be collected after every move (defined in the
algorithm). Statistics whose collection time scales as $\mathcal
O\p{N}$, meaning that doubling $N$ doubles collection time, can be
collected after each iteration. In general, collection with $\mathcal
O\sp{\chi\p{N}}$ time scaling should not occur more often than once
every $\chi\p{N}$ moves (where $\chi\p{N}$ may be an arbitrary
function of only $N$, void of constant prefactors, e.g. $N\log N$, or
$2^N$). These collection rules ensure that scaling up simulations does
not cause them to asymptotically spend all of their time collecting
statistics, and no time actually simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

In this work, we are concerned with simulating the \emph{homogeneous}
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we employ periodic boundary conditions. The use of a
finite cell with periodic boundary conditions suppresses all density
fluctuations on scales larger than the dimensions of the simulated
fluid cell, thereby introducing a source of error. Addressing and
sequestering this error, however, is outside the scope of this work,
and involves considering the limit of numerical results as the number
of spheres $N\to\infty$ (with a constant filling fraction).

@@

To find $\bk{X}_T$, we first need to understand the concept of a
density of states. Given an arbitrary system property $Y\p{s}$
(e.g. energy) determined by the microstate $s$, one can define a
density of states $D\p{Y}$ such that
\begin{enumerate*}[label=\roman*)]
\item $D\p{Y_0}$ is proportional to the number of microstates $s$ for
  which $Y\p{s}=Y_0$, and
\item $\sum_YD\p{Y}=1$.
\end{enumerate*}
The temperature dependence of $\bk{X}_T$ can be expressed in terms of
the expectation value $\bk{X}_E$ and the density of states $D\p{E}$ by
\begin{align}
  \bk{X}_T=\f1{Z\p{T}}\sum_E\bk{X}_ED\p{E}e^{-E/kT},
  \label{eq:XT_norm}
\end{align}
where the partition function $Z\p{T}$ is simply a normalization
factor, given by
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will use a partition function with an unnormalized
density of states,
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  \bk{X}_T=\f1{\tilde Z\p{T}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT}.
  \label{eq:XT}
\end{align}

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos_norm}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A ``histogram method'' is an algorithm for determining a
weight function appropriate for a particular simulation.

@@

The idea behind the simple flat method is such: after simulating for
some time, the energy histogram $H\p{E}$ will have a peak, or some
other uneven distribution of observations, in a region of energies $R$
for which $H\p{E\in R}$ statistically significant. One can thus use
the existing histogram to construct weights which should flatten
$H\p{R}$. Subsequent simulations should then spend less time at
energies with the highest state densities, and more time at energies
with relatively low state densities. The energy histogram from such
simulations should therefore be statistically significant in a larger
range of energies, improving the available estimate for a density of
states with which to compute flat histogram weights. One could then
use the improved estimate to construct new weights, and repeat this
simulation and updating process until simulations satisfy some end
condition (in fact, all histograms will require end conditions, which
is discussed in Section \ref{sec:end_conditions}).

\begin{algorithm}[!b]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

  \item Set $k=0$, set initial weights $w_0\p{E}=1$ for all $E$, and
    choose an initial number of iterations $n_0$ for which to
    simulate.

  \item Simulate for $n_k$ iterations with weights $w_k\p{E}$,
    collecting a histogram $H_k\p{E}$ of energy observations after
    every move.
    \label{alg:simple_flat_sim}

  \item If some predetermined initialization end condition has
    \emph{not} been met, then:
    \begin{enumerate}
    \item increment $k\leftarrow k+1$,
    \item set $n_k=u\p{n_{k-1}}$ with a predetermined function
      $u:u\p{n}\ge n$,
      \label{alg:simple_flat_f}
    \item update weights as $w_k\p{E}=w_{k-1}\p{E}/H_{k-1}\p{E}$ for
      all $E$ at which $H_{k-1}\p{E}\ne0$,
      \label{alg:simple_flat_weight_update}
    \item return to step \ref{alg:simple_flat_sim}.
    \end{enumerate}

@@

Algorithm \ref{alg:simple_flat} provides an implementation of the
simple flat method. This algorithm has two free ``parameters'': the
initial number of iterations $n_0$ for which to simulate, and the
function $u\p{n}$ used in step \ref{alg:simple_flat_f}. With an
appropriate choice of the function $u\p{n}$, this algorithm should not
be particularly sensitive to the value of $n_0$.

The most interesting part of Algorithm \ref{alg:simple_flat} is step
\ref{alg:simple_flat_weight_update}, which updates the weights as
$w_k\p{E}=w_{k-1}\p{E}/H_{k-1}\p{E}$. The reasoning behind this step
is such: say that for two energies $E_0$ and $E_b$ we have that
$H\p{E_0}=\bk{H}_E$ and $H\p{E_b}=b\bk{H}_E$, where $\bk{H}_E$ is the
mean histogram value over all energies. In this case, we wish to
decrease the current simulation bias on $E_b$ by a factor of $b$ while
keeping the bias on $E_0$ the same. As the weights $w\p{E}$ are
proportional to the simulation bias on $E$, dividing the weights
$w\p{E}$ by the current histogram $H\p{E}$ achieves the desired
changes to the simulation biases. The factor $\bk{H}_E$ in this weight
update does not affect results, and is in practice immediately scaled
out of the weight array.

In our implementation of the simple flat method, we chose $u\p{n}=2n$
so that even if the choice of initial $n_0$ is too small, $n_k=2^kn_0$
should quickly approach the appropriate scale for the system at
hand. We also used $n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is
the number of energy levels in the given square-well fluid and
$T_{\t{min}}$ is a minimum temperature of interest for the current
simulation. The factor of $L$ appears because initialization time
should scale with the range of energies of the simulated system. If we
have twice the number of energies, we should explore energy space for
twice as long. To estimate $L$, we multiplied the maximum number of
spheres $M$ which fit within a radius of $\lambda\sigma$ of a single
sphere on a face-centered cubic lattice (i.e. the maximum number of
spheres with which a given sphere could interact) by the total number
of spheres $N$, and divided the result by two so as to not
double-count interactions between spheres (so $L=MN/2$). The
exponential $\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for
two adjacent energies, appears because lower minimum temperatures of
interest warrant sampling lower energies, which in turn requires
simulating for longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (see Section \ref{sec:min_energy}). In this paper,
we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

The Wang-Landau method is the first published histogram method we
discuss, and is the standard method\fixme{cite} to achieve broad
energy sampling in Monte Carlo simulations\cite{wang_landau,
  wang_landau_analysis, wang_landau_mod}.

\fixme{Explain the motivation behind the Wang-Landau method, and
  explain/provide the algorithm.}

\begin{algorithm}[tb]
  \caption{Wang-Landau initialization of weights}
  \label{alg:wang_landau}
  \begin{alg}

@@

  \item Set $k=0$, choose some factor $f_0>0$, and initialize
    $w\p{E}=1$ for all $E$.

@@

  \item If the histogram $H\p{E}$ is sufficiently flat, i.e. if it
    satisfies some predetermined ``flatness'' condition $C\sp{H}$,
    then
    \begin{enumerate}[label=(\alph*)]
    \item increment $k\leftarrow k+1$,
    \item set $f_k=u\p{f_{k-1}}$ with a predetermined function
      $u:u\p{f}<f$,
    \item if $f_k>c$ for some predetermined $c$, return to step
      \ref{alg:wang_landau_sim}.
    \end{enumerate}

@@

>>>>

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased (i.e. Metropolis) Monte Carlo
simulations are incapable of sampling the low energy fluid states
which dominate the partition function at low (i.e. liquid-state)
temperatures. Previous works have developed several generic Monte
Carlo histogram methods for collecting statistics on low energy system
states, but little or no literature exists on these methods'
systematic comparison, as well as their application to the square-well
fluid. The square-well fluid in particular introduces application
challenges not manifest in traditional models for testing and
benchmarking such numerical techniques (e.g. the Ising model). We
propose to implement these methods in Monte Carlo simulations of the
square-well fluid, determine appropriate performance metrics for their
comparison, and identify those methods which are most efficient and
effective at sampling otherwise improbable system states.

@@

\fixme{give more intro about phase transitions and the critical
  point?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the boundary between its liquid and
gaseous phases. The square-well fluid is the simplest system with a
liquid-vapor phase transition, and is therefore of interest for
studying the critical point of fluids. For a generic square-well fluid
to be observed in a liquid state, it must have a localized
distribution of the spheres which make up the fluid. Such a
distribution implies a low energy fluid state, as all interactions
between spheres are attractive. Conversely, a low energy liquid state
of a square-well fluid otherwise observable in both liquid and gaseous
phases implies a localized distribution of spheres. The search for
liquid states of the square-well fluid is thus equivalent to the
search for low energy states.

Monte Carlo simulations are a standard means for studying the
equilibrium thermodynamic properties of a system. In the most
straightforward implementation of Monte Carlo fluid simulations,
however, highly localized distributions of fluid elements are
extremely unlikely to occur, making such simulations impractical for
studying the critical point of the square-well fluid. Several generic
computational methods, called histogram methods, exist for dealing
with such a problem; namely, that interesting regions of a system's
state space cannot be sampled via standard Monte Carlo methods in a
reasonable amount of time. There is no common consensus, however, on
the relative efficacy of these methods. Furthermore, some methods,
such as the Wang-Landau method\cite{wang_landau}, require tuning
several free parameters that lack both ``obvious'' values and
systematic means for assigning them. This work will thus attempt to
quantify and compare the effectiveness of a few such methods as
applied to the square-well fluid.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased, Metropolis Monte
Carlo fluid simulations. Such an algorithm is ``unbiased'' in the
sense that it collects statistics (i.e. data) on all valid system
configurations with equal probability. Statistics whose collection
time scales as $\mathcal O\p{1}$ with system size, meaning that
increasing the number of spheres $N$ in the simulation does not affect
collection time, can be collected after every move (defined in the
algorithm). Statistics whose collection time scales as $\mathcal
O\p{N}$, meaning that doubling $N$ doubles collection time, can be
collected after each iteration. In general, collection with $\mathcal
O\sp{\chi\p{N}}$ time scaling should not occur more often than once
every $\chi\p{N}$ moves (where $\chi\p{N}$ may be an arbitrary
function of $N$ void of constant prefactors, e.g. $N\log N$, or
$2^N$). These collection rules ensure that scaling up simulations does
not cause them to asymptotically spend all computation time collecting
statistics, and no time actually simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

In this work, we are concerned with simulating the \emph{homogeneous}
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we employ periodic boundary conditions. The use of a
finite cell with periodic boundary conditions suppresses all density
fluctuations on scales larger than the dimensions of the simulated
fluid cell, thereby introducing a source of error. Addressing and
sequestering this error, however, is outside the scope of this work,
and involves considering the limit of numerical results as the number
of spheres $N\to\infty$ (keeping the filling fraction $\eta$
constant).

@@

To find $\bk{X}_T$, we first need to understand the concept of a
density of states. Given an arbitrary system property $Y\p{s}$
(e.g. energy) determined by the microstate $s$, one can define a
density of states $D\p{Y}$ such that
\begin{enumerate*}[label=\roman*)]
\item for arbitrary $Y_0$, the value of $D\p{Y_0}$ is proportional to
  the number of microstates $s$ for which $Y\p{s}=Y_0$, and
\item $\sum_YD\p{Y}=1$.
\end{enumerate*}
The units of $D\p{Y}$ are inverse to the units of $Y$. The temperature
dependence of $\bk{X}_T$ can be expressed in terms of the expectation
value $\bk{X}_E$ and the density of states $D\p{E}$ by
\begin{align}
  \bk{X}_T=\f1{Z\p{T}}\sum_E\bk{X}_ED\p{E}e^{-E/kT},
  \label{eq:XT_norm}
\end{align}
where the partition function $Z\p{T}$ is simply a normalization
factor, given by
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will generally use a partition function with an
unnormalized density of states $\tilde D\p{E}$, i.e.
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  \bk{X}_T=\f1{\tilde Z\p{T}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT}.
  \label{eq:XT}
\end{align}

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos_norm}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A histogram method is simply an algorithm or procedure for
determining a weight function appropriate for a particular simulation.

@@

The idea behind the simple flat method is such: after simulating for
some time, the energy histogram $H\p{E}$ will have a peak, or some
other uneven distribution of observations, in a region of energies $R$
for which $H\p{E\in R}$ statistically significant. One can thus use
the existing histogram to construct weights which should flatten
$H\p{R}$. Subsequent simulations should then spend less time at
energies with the highest state densities, and more time at energies
with relatively low state densities. The energy histogram from such
simulations should therefore be statistically significant in a larger
range of energies, improving the available estimate for a density of
states with which to compute flat histogram weights. One could then
use the improved estimate to construct new weights, and repeat this
simulation and updating process until simulations satisfy some end
condition (in fact, most histogram methods will require end
conditions, which is discussed in Section \ref{sec:end_conditions}).

\begin{algorithm}[!t]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

  \item Set $k=0$, choose an initial number of iterations $n$ for
    which to simulate, and initialize weights $w\p{E}=1$ for all $E$.

  \item Simulate for $n$ iterations (with weights), collecting a
    histogram $H\p{E}$ of energy observations after every move.
    \label{alg:simple_flat_sim}

  \item If some predetermined initialization end condition has
    \emph{not} been met, then
    \begin{enumerate}
    \item increment $k\leftarrow k+1$
    \item set $n_k=un_{k-1}$ with a predetermined factor $u\ge1$,
      \label{alg:simple_flat_f}
    \item update weights as $w\p{E}\leftarrow w\p{E}/H\p{E}$ for all
      $E$ at which $H\p{E}\ne0$,
      \label{alg:simple_flat_weight_update}
    \item return to step \ref{alg:simple_flat_sim}.
    \end{enumerate}

@@

Algorithm \ref{alg:simple_flat} provides an implementation of the
simple flat method. This algorithm has two free parameters: the
initial number of iterations $n_0$ for which to simulate and the
factor $u$. Due to the exponential growth of $n_k=u^kn_0$, this
algorithm should not be sensitive to the value of $n_0$.

The most interesting part of Algorithm \ref{alg:simple_flat} is step
\ref{alg:simple_flat_weight_update}, which updates the weights as
$w\p{E}\leftarrow w\p{E}/H\p{E}$. The reasoning behind this step is
such: say that for two energies $E_0$ and $E_b$ we have that
$H\p{E_0}=\bk{H}_E$ and $H\p{E_b}=b\bk{H}_E$, where $\bk{H}_E$ is the
mean histogram value over all energies. In this case, we wish to
decrease the current simulation bias on $E_b$ by a factor of $b$ while
keeping the bias on $E_0$ the same. As the weights $w\p{E}$ are
proportional to the simulation bias on each respective energy $E$,
dividing the weights $w\p{E}$ by the current histogram $H\p{E}$
achieves the desired changes to the simulation biases. The factor
$\bk{H}_E$ in this weight update has no effect on the function of the
weights, and is in practice immediately scaled out of the weight
array.

In our implementation of the simple flat method, we used $u=2$ and
$n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is the number of
energy levels of the given square-well fluid and $T_{\t{min}}$ is a
minimum temperature of interest for the current simulation. The factor
of $L$ appears because initialization time should scale with the range
of energies of the simulated system. If we have twice the number of
energies, we should explore energy space for twice as long. To
estimate $L$, we multiplied the maximum number of spheres $M$ which
fit within a radius of $\lambda\sigma$ of a single sphere on a
face-centered cubic lattice (i.e. the maximum number of spheres with
which a given sphere could interact) by the total number of spheres
$N$, and divided the result by two so as to not double-count
interactions between spheres (so $L=MN/2$). The exponential
$\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for two adjacent
energies, appears because lower minimum temperatures of interest
warrant sampling lower energies, which in turn requires simulating for
longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (addressed in Section \ref{sec:energy_range}). In
this paper, we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_analysis, wang_landau_mod}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations. \fixme{more intro
  on WL}

Unlike the simple flat method, the Wang-Landau method modifies the
weight array on the fly. After every move during initialization
Wang-Landau decreases the weight $w\p{E}$ on the current energy $E$ by
some factor of $f$, so as to decrease weights on the most commonly
observed energies relative to those of the least commonly observed
energies. Eventually, such a simulation with constant modification of
the weights $w\p{E}$ should yield a flat histogram
$H\p{E}\approx\bk{H}_E$ for all $E$, as the modifications to $w\p{E}$
increasingly pushes the simulated system away from the most sampled
energies, and toward the least sampled energies. When the energy
histogram is sufficiently flat, $H\p{E}$ is reset (i.e. set to
$H\p{E}=0$ for all $E$), the factor $f$ is decreased (geometrically
approaching unity), and the entire process is repeated until $f-1$
falls below some cutoff $c\ll 1$.

Resetting and repeating this initialization process is necessary to
perform fine tuning and refinement of the weights. A flat histogram
after a simulation in which $w\p{E}$ has changed does not guarantee
that simulating with the final weights will likewise result in a flat
histogram. Eventually, when $f-1<c\ll 1$, i.e. $f\approx 1$, the
initialization process no longer makes any appreciable modifications
to the weights $w\p{E}$, so that upon achieving a flat histogram one
can be confident that a regular simulation which fixes the current
weights will yield similar results.

Interestingly, the flatness condition $C\sp{H}$ for updating $f$ and
restarting the process should not be too stringent: if the histogram
is completely flat, say $H\p{E}=H_0$ for all $E$, then the weights
$w\p{E}$ have all been modified by the same factor of $f^{H_0}$.
Remembering the scale-invariance of $w\p{E}$, this result would mean
that the weights have not actually changed at all! Similarly, too lax
of a flatness condition will prompt the initialization process to
\begin{enumerate*}[label=\roman*)]
\item reset and decrease $f$ too early, i.e. when the weight array
  could still use some heavy handed adjustment, and
\item quit before one has any reason to think that simulations might
  yield a flat histogram.
\end{enumerate*}
Tuning the flatness condition is thus crucial to the success of the
Wang-Landau method.

\begin{algorithm}[!b]
  \caption{Wang-Landau initialization of weights}
  \label{alg:wang_landau}
  \begin{alg}

@@

  \item Set $k=0$, choose some factor $f_0>1$, and initialize
    $w\p{E}=1$ for all $E$.

@@

  \item If the histogram $H\p{E}$ is not sufficiently flat, i.e. if it
    fails a ``flatness'' condition $C\sp{H}$, return to step
    \ref{alg:wang_landau_sim}, otherwise
    \begin{enumerate}
    \item reset $H\p{E}=0$ for all $E$,
    \item increment $k\leftarrow k+1$,
    \item set $f_k=\sqrt[u]{f_{k-1}}$ for some predetermined $u>1$,
      \label{alg:wang_landau_update}
    \item if $f_k\ge c$ for some predetermined $c$, return to step
      \ref{alg:wang_landau_sim}.
    \end{enumerate}

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
initialization process, which has five free parameters: the number of
iterations $n$ to run in step \ref{alg:wang_landau_sim}, the initial
factor $f_0$, the update factor $u$, the flatness condition $C\sp{H}$,
and the cutoff $c$. The first of these parameters, $n$, controls how
often to check the flatness condition $C\sp{H}$ and run updates, if
necessary. The value of $n$ is not too important, so long as it is
reasonable. Too small value of $n$ will waste computation time
determining $C\sp{H}$, while too large of a value will waste time
simulating when $C\sp{H}$ has already been satisfied. In general, $n$
should be proportional to the computation time of $C\sp{H}$, which
scales with energy range of the simulated system. We therefore use
$n=L$, where $L$ is the number of energy levels for the given
square-well fluid (as discussed in Section \ref{sec:simple_flat}).
Papers in the literature typically neglect providing any heuristic or
formula for $n$, reporting a system-independent $n=10^3$ or
$10^5$\cite{wang_landau, wang_landau_mod}.

The next parameter, $f_0$, controls the initial amount by which to
adjust the weight of the current energy after each move. As mentioned
in Algorithm \ref{alg:wang_landau}, one should always have $f_0>1$,
but the appropriate value of $f_0$ will generally depend on the system
at hand. Too large a value of $f_0$ will magnify stochastic error in
initialization, while too small a value of $f_0$ will cause
simulations to take an exceedingly long time to converge.  Lacking any
heuristics for assigning $f_0$, literature typically suggests
$f_0=e$\cite{wang_landau}; after some experimentation with the free
parameters in the Wang-Landau method, we used $f_0=e^{1/8}$ (in
practice, one usually works with $\ln f$, making the commonly
suggested value $\ln f_0=1$ and our own $\ln f_0=1/8$).

The factor $u$ appearing in step \ref{alg:wang_landau_update} of
Algorithm \ref{alg:wang_landau} controls the amount by which to adjust
$f$ when the histogram $H\p{E}$ is sufficiently flat. In principle,
one need not take a root of $f$ at each reset: the general idea behind
Algorithm \ref{alg:wang_landau} requires only $f_k<f_{k-1}$ to
converge (assuming that $C\sp{H}$ can be satisfied) and
$\lim_{k\to\infty}f_k=1$ to terminate. Taking the $u$-th root of $f$
at each increment of $k$ is a natural and convenient iterative means
to satisfy these requirements. Previous papers have used $u=2$, which
we have adopted as well\cite{wang_landau, wang_landau_mod}.

\fixme{Walk through $C\sp{H}$ and $c$. Finish discussion of WL.}
